---
title: 'The River Ouse Project: Meadow Surveys Data Analysis Annex 2. Results'
author: John Pilkington
date: "2023-06-09"
output:
  html_document: 
     keep_md: FALSE
     
  code_folding: hide
  df_print: paged
  fig.width: 10 #3
  fig.height: 10 #2
     
bibliography: references.bib

---    
<style type="text/css">
   .main-container {max-width: 100%;}
   .row {display: flex;}
   .column {flex: 50%;}
</style>
---
```{css, echo=FALSE}
h1, h2 {text-align: center;}
```

```{r setup, include = FALSE, echo = FALSE, message = FALSE, error = FALSE, warnings = FALSE}
load(".RData")
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(RColorBrewer)
library(knitr)
library(kableExtra)
```

# Annex 2. Analysis by quadrat.

Data tables and figures from the meadow surveys data analysis are presented here with minimal explanation but in a printable form for easy reference. Source code is in Q_SBM_data_analysis.R. The model was built by code in QSBM_build.R. The analysis is quadrat based (i.e. data not aggregated over site), and is Bernouilli SBM with covariates. It is saved in Q_SBM_cov_P.rds

## 1. Latent community membership tables.

These are lists of the plants (prototypes) in each latent community. The count column is the number of times the plant was observed in the survey; the frequency of occurrence in the survey expressed as %.

\pagebreak

```{r echo = FALSE, results = "asis"}
prototypes_list <-  read_rds("lc_prototype_species.rds")
pts_n <- prototypes_list[[1]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[2]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 2 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[3]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 3 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[4]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 4 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[5]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 5 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[6]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 6 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

## 2. Latent community graphs

These graphs may be useful in assessing the structure of the latent communities. Each circle (node) represents a plant, the lines between nodes show which plants form the dyads of the community. The size of the nodes give an idea of the dominance of a plant based on its occurrence frequency (%) and the colour intensity indicates the number of links the plant has in the community. The grey lines represent associative links, the red lines, dissassociative links. 


```{r echo = FALSE, warning=FALSE}
meso_plot_list <-  read_rds("mesoscopic_plots.rds")
plot(meso_plot_list[[1]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[2]])
```

<!-- \pagebreak -->

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[3]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[4]])
```

<!-- \pagebreak -->

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[5]])
```

\pagebreak

NOTE: because latent community 6 is not fully connected, it is not possible to use the same layout as for LC 1 - 5. 

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[6]])
```

\pagebreak

## 3.<a name="section3"></a>Bipartite graphs, community expressions by site and species.

The bipartite graphs that follow show how the species in a community relate to the sites. The size of the site symbols indicates the degree of expression at that site. These graphs can be difficult to interpret, but may give some insight into how community expression works. 

Unfortunately, these figures are too dense to permit easy labelling of each node.

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
bipartite_plot_list <-  read_rds("bipartite_plots.rds")
plot(bipartite_plot_list[[1]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[2]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[3]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[4]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[5]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[6]])
```

\pagebreak

## 4. Polar plot: average LC expression by quadrat at site.

Polar plot relating latent community expression to the survey sites. Sites were sampled with at least five quadrats; at a few sites more quadrats were used. The plot demonstrates the average community expressions of the quadrats at each site.

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
polar_plot <-  read_rds("polar_plot.rds")
plot(polar_plot)
```
