---
title: 'Meadow Surveys Data Analysis Annex: Tables and Figures'
author: John Pilkington
date: "2023-06-09"
output:
  html_document: 
     keep_md: FALSE
     
  code_folding: hide
  df_print: paged
  fig.width: 10 #3
  fig.height: 10 #2
     
bibliography: references.bib

---    
<style type="text/css">
   .main-container {max-width: 100%;}
   .row {display: flex;}
   .column {flex: 50%;}
</style>
---
```{css, echo=FALSE}
h1, h2 {text-align: center;}
```

```{r setup, include = FALSE, echo = FALSE, message = FALSE, error = FALSE, warnings = FALSE}
load(".RData")
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(RColorBrewer)
library(knitr)
library(kableExtra)
```

# Annex I. Analysis by quadrat.

Data tables and figures from the meadow surveys data analysis are presented here with minimal explanation but in a printable form for easy reference. Source code is in Q_SBM_data_analysis.R. The model was built by code in QSBM_build.R. The analysis is quadrat based (i.e. data not aggregated over site), and is Bernouilli SBM with covariates. It is saved in Q_SBM_cov_P.rds

## 1. Latent community membership tables.

These are lists of the plants (prototypes) in each latent community. The count column is the number of times the plant was observed in the survey; the frequency of occurrence in the survey expressed as %.


```{r echo = FALSE, results = "asis"}
prototypes_list <-  read_rds("lc_prototype_species.rds")
pts_n <- prototypes_list[[1]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[2]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[3]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[4]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[5]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

```{r echo = FALSE, results = "asis"}
pts_n <- prototypes_list[[6]] |> activate(nodes) |> as_tibble()
pts_n <- pts_n |> mutate(num(frequency, sigfig = 3)) |>
    select(-frequency, -latent_community) |> rename(frequency = 3)
pts_n |> kbl(caption = "Latent Community 1 species", escape = F) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F, font_size = 12)
```

\pagebreak

## 2. Latent community graphs

These graphs may be useful in assessing the structure of the latent communities. Each circle (node) represents a plant, the lines between nodes show which plants form the dyads of the community. The size of the nodes give an idea of the dominance of a plant based on its occurrence frequency (%) and the colour intensity indicates the number of links the plant has in the community. The grey lines represent associative links, the red lines, dissassociative links. 


```{r echo = FALSE, warning=FALSE}
meso_plot_list <-  read_rds("mesoscopic_plots.rds")
plot(meso_plot_list[[1]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[2]])
```

<!-- \pagebreak -->

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[3]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[4]])
```

<!-- \pagebreak -->

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[5]])
```

\pagebreak

NOTE: because latent community 6 is not fully connected, it is not possible to use the same layout as for LC 1 - 5. 

```{r echo = FALSE, warning=FALSE}
plot(meso_plot_list[[6]])
```

\pagebreak

## 3.<a name="section3"></a>Bipartite graphs, community expressions by site and species.

The bipartite graphs that follow show how the species in a community relate to the sites. The size of the site symbols indicates the degree of expression at that site. These graphs can be difficult to interpret, but may give some insight into how community expression works. For example, in the next graph, for site expressions of latent community 1, there is one isolated species node on the left side. It represents the grass *Lolium perenne*; reference to the species graph for latent community 1 shows that *L. perenne* is the only plant in this community to have dissassociative edges, and in fact all its edges are dissassociative. The sites to the left of this node have zero expression of LC1 - but, obviously, do have *L. perenne*. Nodes to the right of *L. perenne* form a group of sites that do contain *L. perenne*, but nevertheless do otherwise express LC1. The green nodes to the right again are all the other plants in LC1, and the strongest expressions of the community are found in the site nodes to the right again; these sites have a good mix of the non-*L. perenne* species, but not *L. perenne* itself. So a site may express LC1 with or without *L. perenne*, but when *L. perenne* is present, the other species are found less frequently than would be expected by chance (we know this because the *L. perenne* links are dissassociative). This is a clear illustration of the meaning of dissassociative links: it suggests that the dominant grass may be excluding the less frequent members of the community. Some of the LC1-expressing sites have been managed recently to restore species richness; *L. perenne* may be a kind of relic. Alternatively, at some sites former hay meadows may now be managed intensively, and the anomalous presence of the woodland flower *Anemone nemorosa* may indicate that former woodland has been brought into intensive cultivation.

Unfortunately, these figures are too dense to permit easy labelling of each node.

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
bipartite_plot_list <-  read_rds("bipartite_plots.rds")
plot(bipartite_plot_list[[1]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[2]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[3]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[4]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[5]])
```

\pagebreak

```{r echo = FALSE, warning=FALSE, fig.dim = c(10, 10)}
plot(bipartite_plot_list[[6]])
```

\pagebreak


