---
title: 'Stochastic Block Model: a Tool for investigating community structure in Mesotrophic
  Grassland.'
date: "2023-04-21"
output:
  html_document:    
    code_folding: hide
    df_print: paged
  pdf_document: default
fig.width: 6
fig.height: 4
editor_options: 
  markdown: 
    wrap: 72
---
```{r setup, include = FALSE, echo = FALSE}
library("RMySQL")
library(tidyverse)
library(igraph)
library(tidygraph)
library(dplyr)
library(ggraph)
library(sbm)
```

# Part 1: The toolbox. 
## Motivation for Stochastic Block Model (SBM).
Streamside meadows in the Sussex Ouse catchment have been surveyed since
2006 by volunteers from The River Ouse Project, described in detail
[here](http://www.sussex.ac.uk/riverouse/about). The data (to 2019) are
in a database and available for analysis, but consist of more than 24000
records of about 300 plant species from 260 surveys. For this reason, an
unbiased summary avoiding preconceived ideas, arbitrary decisions or
reductionist intent is required. Stochastic Block Model appears to offer
this, in ways to be described later.

## Data wrangling.

Sites were sampled and the plants recorded using at least 5 2m x 2m
quadrats at each survey, as specified for NVC surveys of mesotrophic
grassland (Rodwell ...). Each of the 260 sample sites was surveyed just
once, so in this context, "survey" is equivalent to "site" or
"location". There is no replication in the data set, and we treat it as
a synoptic view of meadows in the river Ouse catchment, ignoring any
temporal drift over the study period; as though all 1457 quadrats were
thrown on the same day.

We are interested in plants that occur together more or less frequently
than would be expected by chance. Pairwise contingency tables for every
combination of the plants that we encountered were assessed with
Fisher's exact test, returning a p_value for the significance of any
association, and also the odds-ratio, an estimate of the strength and
sign of the association (inclusive or exclusive). Plant pairs (diads)
for which the p-value was less than 0.05 were accepted for further
analysis. Arguably, this is an arbitrary decision; but it seems
conventional enough. Further, some diads with a low p-value had the
odds-ratio flagged as "Inf"; after some examination, these also were
excluded from further analysis.

The p-values and logarithms of the odds-ratios (lor) were retained, but
it should be understood that *the SBM itself has no knowledge of these
quantities*. They are merely intended to be useful for onwards
investigation. The only thing the SBM knows about is whether or not an
association was detected for each dyad.

Code for data wrangling is shown here.
```{r}
d <- read.csv("hits.csv")
# What follows thanks to Brian Shalloway: 
# https://www.bryanshalloway.com/2020/06/03/tidy-2-way-column-combinations/#fn4
# Manipulating the data into contingency tables for species pairs (dyads)
# and getting the fisher.test statisics using map not for-loops.
df_lists <- d %>%
  summarise_all(list) %>% 
  pivot_longer(cols = everything(), 
               names_to = "var", 
               values_to = "vector")

df_lists_comb <- expand(df_lists,
                        nesting(var, vector),
                        nesting(var2 = var, vector2 = vector))

df_lists_comb_as <- df_lists_comb %>% 
  filter(var != var2) %>% 
  arrange(var, var2) %>% 
  mutate(vars = paste0(var, ".", var2)) %>% 
  select(contains("var"), everything())

is_event <- function(...){ # See map2_int below
  sum(..1 & ..2)
}

df_lists_comb_as <- df_lists_comb %>% 
  filter(var != var2) %>% 
  mutate(events = map2_int(.x = vector, .y = vector2, .f = is_event)) %>% 
  filter((events > 0) & events < length(df_lists$vector[[1]])) %>%
  select(-events)
rm(is_event)

# function to make the contingency tables.
xtab <- function(...){ # See map2 below
  t <- table(..1, ..2) 
}
# And map them into a convenient data structure.
df_tables <- df_lists_comb_as %>% 
  mutate(xtable = map2(vector, vector2, xtab))
# Clean up...
rm(df_lists, df_lists_comb, df_lists_comb_as, xtab)

# Make a graph and simplify it before doing the fisher test
df_edges <- df_tables %>% select(var, var2) %>% rename(from = var, to = var2)
g1 <- as_tbl_graph(graph.data.frame(d = df_edges, directed = FALSE))

# Add edge attributes to g1 edges
g1 <- g1 %>% activate("edges") %>%
  mutate(xtable = df_tables$xtable)

# Simplify the Tidygraph way
g1 <- g1 %>% activate("edges") %>% 
  filter(!edge_is_multiple()) %>%
  filter(!edge_is_loop())
# Clean up ...
rm(df_edges, df_tables)

```

## Model building.








The groups that SBM finds exist in the data, they are not necessarily
expressed in the field.
